---
title: "FA1Lec3_DMAW"
author: "MORILLO, JADE MARCO S."
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r }
filtered_diamonds <- diamonds %>%
  filter(carat > 1.7, price < 5000)

head(filtered_diamonds)
```

It can be observed that there are only a single available diamond that is at least 1.7 carats and priced under 5000.

```{r }
selected_diamonds <- diamonds %>%
  select(carat, cut, price)

head(selected_diamonds)
```

This shows three of the most useful variables when trading diamonds, helping both sellers and buyers to easily negotiate.

```{r }
arranged_diamonds <- diamonds %>%
  arrange(desc(price))

head(arranged_diamonds)
```

It is revealed that the most expensive  is priced at 18,823 which is a premium cut 2.29-carat diamond.

```{r }
mutated_diamonds <- diamonds %>%
  mutate(price_per_carat = price / carat)

head(mutated_diamonds)
```

This added a new column that reveals the price per carat of each diamond with the first one priced at 1,417 per carat.

```{r }
summary_stats <- diamonds %>%
  summarize(mean_price = mean(price),
            max_price = max(price),
            min_price = min(price))

summary_stats
```

The mean price of all the diamonds is 3,933, with the most expensive being priced at 18,823 while the cheapest is 326.

```{r }
grouped_summary <- diamonds %>%
  group_by(cut) %>%
  summarize(mean_price = mean(price))

grouped_summary
```

This shows that the mean price of premium is the highest among other cuts, while the ideal being the lowest.