---
title: "FA1Lec2_DMAW"
author: "MORILLO, JADE MARCO S."
date: "2025-02-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r }
data("diamonds")
head(diamonds)
```

```{r }
ggplot(diamonds, aes(x = cut)) +
  geom_bar() +
  labs(title = "Distribution of Diamond Cuts",
       x = "Cut",
       y = "Count")
```

The bar graph shows that ideal was the most common cut of diamond while fair cut was the least popular.

```{r }
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.1) +
  labs(title = "Distribution of Carat",
       x = "Carat",
       y = "Count")
```

The histogram reveals that the carats are mostly between 0.2 to 1.0, with most being 0.3 carat.

```{r }
ggplot(diamonds, aes(x = x, y = y)) +
  geom_point(alpha = 0.05, color = "purple") +  # Lower alpha for better visibility
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Scatter Plot of Diamond Dimensions (All Data)",
       x = "Length (x)",
       y = "Width (y)")
```

Most diamond have similar length and width as seen by the red line, representing the 'x=y' line.

```{r }
ggplot(diamonds, aes(x = cut, y = price)) +
  geom_boxplot() +
  labs(title = "Price Distribution by Cut",
       x = "Cut",
       y = "Price")
```

Surprisingly, the better the cut of the diamonds are, the lower the median price is, with the premium cut being an exception as it is priced around the same as the fair cut.

```{r }
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.1) +
  facet_wrap(~ cut) +
  labs(title = "Price vs. Carat by Cut",
       x = "Carat",
       y = "Price")
```

This plot shows a better picture of how diamonds are priced as it is directly proportional with the carat of the diamond.